---
title: "Assignment 4"
author: "Abhishek Kumar"
date: "18/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r installpackage}
install.packages("readxl")
```


```{r}
setwd("/home/abhishek/Desktop/R assignment/Assignment 4")
```

## Data Manipulation

#1 Read the data into R.

```{r loaddata} 
library("readxl")
clgoals = read_excel("CLGoals.xlsx")
head(clgoals)
str(clgoals)
```

#2. Produce a table that tabulates frequency of each number of goals.

```{r tablegoals}

table_goals = table(clgoals$Goals)
table_goals # Frequency of each number of goals

```

#3. Produce a plot of the frequency of each number of goals.

```{r freqplot}
barplot(table_goals, xlab ="Number of goals",ylab = "Frequency", main =" Frequency of each number of goals")

```

#4. Calculate the mean and the standard deviation of the number of goals.

```{r meansd}
mean_goals = mean(clgoals$Goals) # Mean goals
sd_goals = sd(clgoals$Goals) # Standard deviation of number of goals
mean_goals
sd_goals 
```



## Task 2a: Poisson Modelling

#1. Write a function that calculates the log-likelihood function (for a specified value of λ) for the Poisson model for the UEFA Champions League data.

```{r loglikelihood}

log_likelihood = function(observations, lambda){
  
  prob_at_X_xi = dpois(observations, lambda = lambda) 
  likelihood = sum(log(prob_at_X_xi))
  return(likelihood)
  
}

log_likelihood(clgoals$Goals, 3)

```

#2. Plot the log-likelihood function for a range of values of λ.

```{r plot}

lambda_range =  seq(from = min(clgoals$Goals), to = max(clgoals$Goals), by =0.1)
log_likelihood_per_lambda = c()

for(i in lambda_range){
  
  log_likelihood_per_lambda = c(log_likelihood_per_lambda, log_likelihood(clgoals$Goals, i))
  
}

log_likelihood_per_lambda # likelihood corresponding to different lambdas
plot(lambda_range, log_likelihood_per_lambda,main = "Rate vs log-Likelihood", type="l", pch=15, col ="blue")
```

  
#3. Add a vertical line to the plot at the value x and visually verify that this maximizes the log-likelihood function.

Maximum likelihood estimates finds the unknown parameter lambda value that maximizes likelihood.
From the graph I can see that log likelihood is maximum at lambda value which was mean value of goals scored in each match, i,e at 3.25.

```{r}
max(log_likelihood_per_lambda)
log_likelihood(clgoals$Goals,mean(clgoals$Goals))
plot(lambda_range, log_likelihood_per_lambda,main = "Rate vs log-Likelihood", type="l", pch=15, col ="blue")
abline(v = lambda_range[which.max(log_likelihood_per_lambda)], col= "red")


```
  
  
  
#4. Simulate 48 values from a Poisson model with λ = x and summarize the resulting values (contrasting them with the summaries produced in Task 1).

Simulating 48 random variables from poisson distribution with lambda as mean of goals scored.

```{r posi}
set.seed(100)
pois_mean_lambda = rpois(48, lambda= mean(clgoals$Goals))
barplot(table(pois_mean_lambda), xlab ="Number of goals",ylab = "Frequency", main =" Frequency of each number of goals")
mean_pois_mean_lambda = mean(pois_mean_lambda)
sd_pois_mean_lambda = sd(pois_mean_lambda)

print(paste("Average goals : ", mean_pois_mean_lambda))
print(paste("Standard deviation : ", sd_pois_mean_lambda))
```


#5. Simulate 48 values from a Poisson model for other values of λ and summarize the resulting values (contrasting them with the summaries produced in Task 1).

```{r}

for(lambda in lambda_range){
  
  pois = rpois(48, lambda = lambda)
  mean_pois_mean_lambda = mean(pois)
  sd_pois_mean_lambda = sd(pois)
  print(table(pois))
  print(paste("Average goals_lambda ", lambda, " :", mean_pois_mean_lambda))
  print(paste("Standard deviation_lambda ", lambda, " :",sd_pois_mean_lambda))
}

barplot(table(pois), xlab ="Number of goals",ylab = "Frequency", main =" Frequency of each number of goals")

```


## Task 2b: Hurdle Modelling

#1. Create a dHurdle() function that has arguments x, param that computes P {X = x} for the Hurdle model, where the first element of the vector param is θ and the second element of the vector param is λ. Ensure that the function can handle x being a vector of values.


```{r hHurdle}

# Function takes in input two vectors x (observations) and param (theta and lambda)
# Calculates the probabilities based on values of x for the hurdle model 
# Returns the vector of Probability of each observations (P_X_x) for a particular theta and lambda values.

dHurdle = function(x, param){
  theta = param[1]
  lambda = param[2]
  P_X_x = c()
  l = length(x)
  for(i in 1:l){
      
      if(x[i] == 0){
       P = theta
       P_X_x = c(P_X_x, P)
     } #if
      else{
       P = (1-theta)*((lambda^x[i])*(exp(-lambda)))/(factorial(x[i])*(1-exp(-lambda)))
       P_X_x = c(P_X_x, P)
     }#else
    }# for
   return(P_X_x)
  
} # function

dHurdle(clgoals$Goals,c(2,3))

```

#2. Write a function that calculates the log-likelihood function (for a specified value of param) for the Hurdle model for the UEFA Champions League goal data.


```{r}

# This functions take in input a vextor of values for which likelihood has to be calculated  and another vector of parameter theta and lambda
# call dHurdle() function to calculate probability of each observations for hurdle model.
# Returns log-likelihood for the hurdle model for a particular set of observations.

log_likelihood_hurdle = function(observations, param){
  
  theta = param[1]
  lambda = param[2]
  
  prob_at_X_xi_hurdle = dHurdle(observations,c(theta,lambda))
  likelihood_hurdle = sum(log(prob_at_X_xi_hurdle))
  return(likelihood_hurdle)
}
log_likelihood_hurdle(clgoals$Goals, c(0.4,4)) # log-likelihood for the hurdle model for the UEFA champions league goal data.

```

#3. Use the optim function to find the value of θ and λ that maximizes the log-likelihood. Hint: optim minimizes functions, by default, so you may want to write a function that computes minus the log-likelihood and minimize that. Alternatively, you can set control=list(fnscale=-1) as an argument in optim to make it maximize.


So after optimizing the log_likelihood_hurdle function for theta and lambda, the results show that we get the maximum likelihood for for the UEFA Champions League goal data at theta = 0.04173414 and lambda = 3.26110189 and the value is -90.30556.

```{r}

optimized_pars = optim(par = c(0.1,0.1) , fn = log_likelihood_hurdle, observations= clgoals$Goals, control=list(fnscale=-1))
optimized_pars
```


#4. Comment on the value of θ found and compare the log-likelihood values found for the Poisson and Hurdle models.

So after optimizing the log_likelihood_hurdle function for theta and lambda, the results show that we get the maximum likelihood for for the UEFA Champions League goal data at theta = 0.04173414, and the value is -90.30556.

For poisson distribution maximum likelihood is achieved approximately at lambda value which is equal to mean of goals scored i.e 3.25 and maximum_likelihood was found to be  -90.32991.

Maximum log likelihood is quite similar in both the distributions with difference of 0.02435 which is very small.
